Intersection Types, Sub-formula Property, and the Functional Character of the Lambda Calculus

### 2018-08-08 DenotSem/determ.tex
progress

\newcommand{\lam}[1]{\lambda #1.\,}
\newcommand{\app}[0]{\;}
\newcommand{\IF}[0]{\mathrm{if}\;}
\newcommand{\THEN}[0]{\;\mathrm{then}\;}
\newcommand{\ELSE}[0]{\;\mathrm{else}\;}
\newcommand{\ESEM}[1]{\mathcal{E}[\![ #1 ]\!]}
\newcommand{\FSEM}[1]{\mathcal{F}[\![ #1 ]\!]}
\newcommand{\SET}[1]{\mathcal{P}(#1)}
\newcommand{\WF}[1]{\mathsf{wf}(#1)}
\newcommand{\UP}[1]{\mathord{\uparrow} #1}

@@

\noindent \textbf{Lemma} (Collapse Duplicates) \\ If
$\Gamma_1,A,A,\Gamma_2 \vdash C$, then $\Gamma_1,A,\Gamma_2 \vdash
C$. \\
%
(This is proved by well-founded induction on the lexicographical
ordering of the pair $(n,k)$ where $n$ is the size of $A$ and $k$ is
the depth of the derivation of $\Gamma_1,A,A,\Gamma_2 \vdash C$. Proof
assistants such as Isabelle and Coq do not directly provide the depth
of a derivation, but the depth can be manually encoded as an extra
argument of the relation, as in $\Gamma_1,A,A,\Gamma_2 \vdash_k C$.) \\

@@

We also introduce the following notation for all the well-formed types
that are super-types of a given type.
\[
   \UP{A} \stackrel{def}{=} \{ B\mid A <: B \text{ and } \WF{B} \}
\]

We shall represent variables with de Bruijn indices, so an environment
$\rho$ is a sequence of types.  The denotational semantics of the CBV
lambda calculus is defined as follows.
\begin{align*}
  \ESEM{n}\rho &= \UP{n} \\
  \ESEM{x}\rho &=  \IF x < |\rho| \THEN \UP{\rho_k} \ELSE \emptyset \\
  \ESEM{\lam{x}e }\rho &= \{ A \mid \WF{A} \text{ and } \FSEM{A}(e,\rho) \} \\
  \ESEM{e_1\app e_2}\rho &= \{ C \mid
    A \in \ESEM{e_1}\rho,
    B \in \ESEM{e_2}\rho,
    A <: B \to C, \text{ and } \WF{C} \} \\
  \ESEM{f(e_1,e_2)}\rho &=
    \{ \}
\end{align*}

>>>>

\newcommand{\lam}[1]{\lambda #1.\,}
\newcommand{\app}[0]{\;}
\newcommand{\IF}[0]{\mathrm{if}\;}
\newcommand{\THEN}[0]{\;\mathrm{then}\;}
\newcommand{\ELSE}[0]{\;\mathrm{else}\;}
\newcommand{\SEM}[1]{[\![ #1 ]\!]}
\newcommand{\ESEM}[1]{\mathcal{E}\SEM{#1}}
\newcommand{\FSEM}[1]{\mathcal{F}\SEM{#1}}
\newcommand{\SET}[1]{\mathcal{P}(#1)}
\newcommand{\WF}[1]{\mathsf{wf}(#1)}
\newcommand{\UP}[1]{\mathord{\uparrow} #1}
\newcommand{\ATOMS}[1]{\mathit{atoms}(#1)}

@@

\noindent \textbf{Lemma} (Collapse Duplicates) \\
%
If $\Gamma_1,A,A,\Gamma_2 \vdash C$, then $\Gamma_1,A,\Gamma_2 \vdash
C$. \\
%
(This is proved by well-founded induction on the lexicographical
ordering of the pair $(n,k)$ where $n$ is the size of $A$ and $k$ is
the depth of the derivation of $\Gamma_1,A,A,\Gamma_2 \vdash C$. Proof
assistants such as Isabelle and Coq do not directly provide the depth
of a derivation, but the depth can be manually encoded as an extra
argument of the relation, as in $\Gamma_1,A,A,\Gamma_2 \vdash_k C$.) \\

@@

Here are some examples and non-examples of well-formed types.
\begin{gather*}
  \WF{4} \qquad \WF{3 \sqcap 3} \qquad \neg \WF{3 \sqcap 4} \\
  \WF{(0\to 1) \sqcap (2 \to 3)} \qquad \neg \WF{(0 \to 1) \sqcap (0 \to 2)}
\end{gather*}
It is sometimes helpful to think of well-formed types in terms of the
equivalence classes determined by subtype equivalence:
\[
A \approx B \quad = \quad
  A <: B \text{ and } B <: A
\]
For example, we have $3 \approx (3 \sqcap 3)$, so they are
in the same equivalence class and $3$ would be the representative.

We also introduce the following notation for all the well-formed types
that are super-types of a given type.
\[
   \UP{A} \quad = \quad \{ B\mid A <: B \text{ and } \WF{B} \}
\]

We shall represent variables with de Bruijn indices, so an environment
$\rho$ is a sequence of types.  The denotational semantics of the CBV
$\lambda$-calculus is defined as follows.
\begin{align*}
  \ESEM{n}\rho &= \UP{n} \\
  \ESEM{x}\rho &=  \IF x < |\rho| \THEN \UP{\rho_k} \ELSE \emptyset \\
  \ESEM{\lambda e }\rho &= \{ A \mid \WF{A} \text{ and } \FSEM{A}(e,\rho) \} \\
  \ESEM{e_1\app e_2}\rho &= \left\{ C\, \middle| \exists A,B.\;
      \begin{array}{l}
      A \in \ESEM{e_1}\rho,
      B \in \ESEM{e_2}\rho,\\
      A <: B \to C, \text{ and } \WF{C}
      \end{array}
        \right\} \\
  \ESEM{f(e_1,e_2)}\rho &=
      \left\{ C\, \middle| \begin{array}{l}
       \exists A,B,n_1,n_2.\; A \in \ESEM{e_1}\rho, B \in \ESEM{e_2}\rho,\\
      \qquad A <: n_1, B <: n_2, \SEM{f}(n_1,n_2) <: C, \WF{C} 
      \end{array} \right\} \\
  \ESEM{\IF e_1 \THEN e_2 \ELSE e_3}\rho &=
    \left\{ B\, \middle| \exists A.\;
    \begin{array}{l}A \in \ESEM{e_1}\rho, A <: n,\\
           n = 0 \Rightarrow B \in \ESEM{e_3}\rho,\\
           n \neq 0 \Rightarrow B \in \ESEM{e_2}\rho
    \end{array}
    \right\}
    \\[2ex]
    \intertext{where}
   \FSEM{n}e\rho &= \mathit{false} \\
   \FSEM{A \sqcap B}e \rho &= \FSEM{A}e\rho \text{ and } \FSEM{B}e\rho\\
   \FSEM{A \to B}e \rho &= B \in \ESEM{e} (A, \rho)
\end{align*}

\noindent \textbf{Proposition} (Consistency is Symmetric and Reflexive) \\
%
\begin{enumerate}
\item If $A \sim B$, then $B \sim A$.
\item If $\WF{A}$, then $A \sim A$.
\end{enumerate}

It will often be convenient to decompose a type into its set of
\emph{atoms}, defined as follows.
\begin{align*}
  \ATOMS{n} &= \{ n \} \\
  \ATOMS{A\to B} &= \{ A \to B \} \\
  \ATOMS{A \sqcap B} &= \ATOMS{A} \cup \ATOMS{B}
\end{align*}

The consistency of two types is determined by the consistency of its
atoms. \\

\noindent \textbf{Lemma} (Atomic Consistency) 
\begin{enumerate}
\item If $A \sim B$, $C \in \ATOMS{A}$, and $D \in \ATOMS{B}$,
  then $C \sim D$.
\item If (for any $C \in \ATOMS{A}$ and $D \in \ATOMS{B}$, 
  $C \sim D$), then $A \sim B$.
\item If $A \not\sim B$, then $C \not\sim D$ for some
  $C \in \ATOMS{A}$ and $D \in \ATOMS{B}$.
\item If $C \not\sim D$, $C \in \ATOMS{A}$, and $D \in \ATOMS{B}$,
  then $A \not\sim B$. 
\end{enumerate}

There are also several properties of subtyping and the atoms of a type.\\

\noindent \textbf{Lemma} (Atomic Subtyping)
\begin{enumerate}
\item If $A <: B$ and $C \in \ATOMS{B}$, then $A <: C$.
\item If $A <: n$, then $n \in \ATOMS{A}$.
\item $n <: A$ if and only if $\ATOMS{A} \subseteq \{ n \}$.
\item If $C <: A \to B$, then $D\to E \in \ATOMS{C}$ for some $D,E$.
\item If $\Gamma \vdash A$ and every atom in $\Gamma$ is a function type,
  then every atom of $A$ is a function type.
\end{enumerate}

And we have the following important inversion lemma for function
types. We use the following abbreviations:
\begin{align*}
  \mathrm{dom}(\Gamma) &= \{ A \mid \exists B.\; A \to B \in \Gamma \}\\
  \mathrm{cod}(\Gamma) &= \{ B \mid \exists A.\; A \to B \in \Gamma \}
\end{align*}

\noindent \textbf{Lemma} (Subtyping Inversion for Funcion Types)\\
%
If $C <: A \to B$, then there is a sequence of function types $\Gamma$
such that
\begin{enumerate}
\item each element of $\Gamma$ is an atom of $C$,
\item For every $D\to E \in \Gamma$, we have $A <: D$, and
\item $\bigsqcap \mathrm{cod}(\Gamma) <: B$.
\end{enumerate}
Note that item 2 above implies that $A <: \bigsqcap \mathrm{dom}(\Gamma)$.

With the above lemmas, we are ready to prove the Consistency and
Subtyping Lemma that previously stumped me.\\

\noindent \textbf{Lemma} (Consistency and Subtyping)
\begin{enumerate}
\item  If $A \sim B$, $A <: C$, and $B <: D$,
  then $C \sim D$.
\item If $A \not\sim B$, $C <: A$, $D <: B$, then $C \not\sim D$.
\end{enumerate}

(1) The proof is by strong induction on the sum of the depths of $A$,
$B$, $C$, and $D$. To show that $C \sim D$ it suffices to show that
all of their atoms are consistent. Suppose $C' \in \ATOMS{C}$
and $D'\in\ATOMS{D}$. So we need to show that $C' \sim D'$.
%Note that $A <: C'$ and $B <: D'$
We proceed by cases on $C'$.
\begin{itemize}
\item Case $C'=n_1$:\\
  We have $A <: C'$ and therefore $n_1 \in \ATOMS{A}$.
  Then because $A \sim B$, we have $\ATOMS{B} \subseteq \{n_1\}$.
  We have $B <: D'$, so we also have $\ATOMS{D} \subseteq \{n_1\}$.
  Therefore $C' \sim D'$.

\item Case $C'=C_1\to C_2$:\\
  We have $A <: C_1 \to C_2$, so by inversion we have
  some sequence of function types $\Gamma_1$ such that
  every element of $\Gamma_1$ is an atom of $A$,
  $C_1 <: \bigsqcap \mathrm{dom}(\Gamma_1)$,
  and $\bigsqcap \mathrm{cod}(\Gamma_1) <: C_2$.

  We also know that $D'$ is a function type, say $D'=D_1 \to D_2$.
  (This is because we have $A <: C'$, so we know that $A_1\to A_2 \in
  \ATOMS{A}$ for some $A_1,A_2$. Then because $A \sim B$, we know that
  all the atoms in $B$ are function types.  Then because $B <: D$ and
  $D' \in \ATOMS{D}$, we have that $D'$ is a function type.)
  %
  So by inversion on $B <: D_1 \to D_2$, we have
  some sequence of function types $\Gamma_2$ such that
  every element of $\Gamma_2$ is an atom of $B$,
  $D_1 <: \bigsqcap \mathrm{dom}(\Gamma_2)$,
  and $\bigsqcap \mathrm{cod}(\Gamma_2) <: D_2$.

  It's the case that either $C_1 \sim D_1$ or $C_1 \not\sim D_1$.
  \begin{itemize}
  \item Sub-case $C_1 \sim D_1$.\\
    It suffices to show that $C_2 \sim D_2$.
    By the induction hypothesis, we have
    $\bigsqcap \mathrm{dom}(\Gamma_1) \sim \bigsqcap \mathrm{dom}(\Gamma_2)$.

    As an intermediate step, we shall prove that
    $\bigsqcap \mathrm{cod}(\Gamma_1) \sim \bigsqcap \mathrm{cod}(\Gamma_2)$,
    which we shall do by showing that all their atoms are consistent.
    Suppose $A' \in \ATOMS{\bigsqcap \mathrm{cod}(\Gamma_1)}$
    and $B' \in \ATOMS{\bigsqcap \mathrm{cod}(\Gamma_2)}$.
    There is some $A_1\to A_2 \in \Gamma_1$ where $A' \in \ATOMS{A_2}$.
    Similarly, there is $B_1 \to B_2 \in \Gamma_2$ where $B' \in \ATOMS{B_2}$.
    Also, we have $A_1 \to A_2 \in \ATOMS{A}$ and $B_1 \to B_2 \in \ATOMS{B}$.
    Then because $A \sim B$, we have $A_1 \to A_2 \sim B_1 \to B_2$.
    Furthermore, we have $A_1 \sim B_1$ because
    $\bigsqcap \mathrm{dom}(\Gamma_1) \sim \bigsqcap \mathrm{dom}(\Gamma_2)$,
    so it must be the case that $A_2 \sim B_2$.
    Then because $A' \in \ATOMS{A_2}$ and $B' \in \ATOMS{B_2}$, we
    have $A' \sim B'$. Thus concludes this intermediate step.

    By another use of the induction hypothesis, we have
    $C_2 \sim D_2$, and this case is finished.

  \item Sub-case $C_1 \not\sim D_1$.\\
    Then we immediately have $C_1 \to C_2 \sim D_1 \to D_2$.
  \end{itemize}

\item Case $C'=C_1\sqcap C_2$:\\
  We already know that $C'$ is an atom, so we have a contradiction
  and this case is vacously true.
\end{itemize}

Now we can move on to lemmas about the denotational semantics that
will be needed to prove the Join Theorem. \\

\noindent \textbf{Lemma} (Weakening)
\begin{enumerate}
\item If $\FSEM{A}e \rho_1$, $\rho_1 <: \rho_2$ and
  $(\forall B \rho_1 \rho_2,\; B \in \ESEM{e}\rho_1, \rho_2 <: \rho_1
  \Rightarrow B \in \ESEM{e}\rho_2)$, then
  $\FSEM{A}e \rho_2$.
\item If $A \in \ESEM{e}\rho_1$ and $\rho_2 <: \rho_1$,
  then $A \in \ESEM{e}\rho_2$.
\end{enumerate}
(Part 1 is proved by induction on $A$. Part 2 is proved by induction
on $e$ and uses part 1.)

\noindent \textbf{Lemma} (Consistent Application) \\
%
If $A_1 \sim A_2$, $B_1 \sim B_2$, $A_1 <: B_1 \to C_1$,
$A_2 <: B_2 \to C_2$, and all these types are well formed,
then $C_1 \sim C_2$.
